Sophize Mathematics Library

### 2022-05-30 math_library/sophize_math.tex
CICM'22 2nd submission

\newcommand\todo[1]{\textcolor{red}{TODO: #1}}
\graphicspath {{../res}}
\begin{document}
%
\title{Sophize Mathematics Library}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Abhishek Chugh\inst{1}\orcidID{0000-0001-6765-5229}}
%
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Sophize Foundation, Bengaluru - 560066, India 
\email{abc@sophize.org}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Sophize is a novel mathematics library and discussion platform with a mission to help our
users find and organize mathematical proofs. We present Sophize's novel yet familiar
knowledge organization scheme that is used to represent a wide variety of proofs. With this
scheme at its core, we have engineered a platform to logically aggregate knowledge from
multiple sources. The platform curates knowledge from published research, encyclopaedias
such as PlanetMath and Wikipedia, informal computer programs, and formal systems such as
Metamath; and allows its users to run a federated search over them. To make this happen, two
developments were necessary. First, we came up with the concept of 'proof-generating
machines' and developed an open plugin architecture that allows proofs from computer
programs to be generated as required. Second, we extended the Markdown language to represent
the connections between mathematical objects found across various sources of knowledge. In
addition, we also utilized this new language to create a novel communication system built
specifically to aid mathematicians in solving problems collaboratively.

@@

In the past couple of decades, the Internet has revolutionized our life by giving us access
to seemingly unlimited information. Information from a wide variety of topics ranging from
human medicine to rocket science is now just a few keystrokes away. However, it seems
that we are having even more trouble making sense of the World around us. In fact, dealing with the vast amounts of information thrown at us seems to be one of the most important problems of our generation. While we have access to a wide range of opinions on almost every topic imaginable, it has become increasingly difficult to decide which one to trust. It
seems that we can find "proofs" for anything and everything we would like to believe.

Thus the author's primary motivation for this project is to create a library that not just
lays out all the information about the World but organizes it in a way that helps us make
better sense of it. A library that helps us access justifications from a variety of
viewpoints and allows us to collaboratively and objectively evaluate issues in the
justifications or the viewpoints they arise from. Clearly, this is a challenging goal.

@@

Logical organization of mathematics can lead to many advantages. We list some interesting ones below.

@@

Another use case would be to understand the consequences of important conjectures like the Reimann hypothesis by creating two belief sets - one where the conjecture is true and one where it is false. The proof graphs would systematically store such conjectural knowledge which often gets lost especially when results from multiple conjectures are involved. Complexity theory is one such field where this need is quite apparent. On a more practical note, several information security systems build on the conjectural developed with the assumption $P \neq NP$.

Proof graphs are currently organized as trees as seen in Fig \ref{proof_tree}. Users can keep exploring the proof till they reach a point where the propositions are already familiar to them. A future application of proof graphs would be to generate narrative proofs based on tailored to the reader's current understanding. This could be achieved in a classroom setting where the current level of mathematical understanding of students is known or assumed. It may also be possible to infer this from the user's activity on the platform.

@@

\section{Proof-Generating Machines (PGM)}
\label{sec:pgm}
A lot of mathematical knowledge is computed. Computation machines (e.g. calculators, algebras) can be seen as performing the following tasks
\begin{itemize}
\item Parsing the input into a language that can represent math formulae. 
\item Finding and outputting an equivalent form of the formula.
\end{itemize}
Thus, their focus is not on finding the proof but only finding the equivalent (simplified) representation of the input formula (Although many computer algebras do show the steps of the computation.). With Sophize, the focus instead is to generate proofs of the input propositions. A \textbf{proof-generating machine} (PGM) is a computer program that generates a partial proof graph against or in favour of an argument. PGMs can also have the capability to perform tasks 1 and 2, mentioned above.

PGMs take in some text as input which is parsed by the each PGM as they wish. Typically the text is treated either as a proposition (e.g. `3 + 5 = 8') or a formula (eg. `5+6') and in the latter case the PGM is responsible for finding an appropriate proposition (eg. `5 + 6 = 11') for the given input. The PGM then generates a proof graph for this proposition and returns back to the user.

@@

\begin{enumerate}
\item Arguments whose the premises are limited to the premises of the PGM.
\item Special `Arguments' that indicate that the generation of the proof of this argument's conclusion must be delegated to another PGM. To paginate large proofs, a PGM may also delegate to itself after generating some (say 100) arguments.
\end{enumerate}
\subsection{Materializing results}

@@

In the above example, the `EXPAND' option expanded the definition of the term, and statement of the proposition in place. Thus we were able to easily re-use concepts and theorems already extracted from other sources (say, Wikipedia and the Oxford dictionary in this case). Clicking on the underlined concepts, pops-up a modal dialog with the definition of the term clicked.

@@

Sophize divides its data into groups called datasets for copyright and data access management. For example, the 'wiki' dataset that extracts knowledge from Wikipedia mathematics glossaries releases its data under `CC-BY-SA' license, whereas the `metamath' dataset is public (`CC0'). Our signed-in users are free to create their own datasets and we allow them to choose who can read, write or comment in these datasets. Another interesting access control is the restriction of use of resources (especially argument-validity criteria) from an otherwise open-to-read dataset.

Information can be added to Sophize platform using intuitive web forms. However, it is infeasible to extract information from large repositories in this manner. Sophize publishes its data schema (explained in Section \ref{sec:core}) and helper libraries in multiple languages. The extracted information is can thus be stored as text (JSON) files in a repository. Sophize provides dataset owners the ability to sync knowledge such repositories to the central Sophize database.

\subsection{Search Interface}
\label{sec:search}
Sophize allows its users to run a run a federated search over its resources using an Elasticsearch server. We even allow users to \LaTeX to their search queries. The server runs search a space-insensitive and case-insensitive search over \LaTeX\space formulae by encoding \LaTeX\space formulae as lexemes and an Elasticsearch analyzer that removes spaces and lower cases the encoding. (See section \ref{sec:sources} for sources)

@@

We show how PGMs can be used to compute proofs in real-time and embed them in our knowledge graphs. We utilize the proof-graphs using Sophize Markdown to allow users to effortlessly browse proofs and lookup math entities in math literature. While there are other ways to represent semantic knowledge, such as content MathML and OpenMath, their scope is somewhat limited to specifying the meaning of the mathematical formula. The sTeX system \cite{Kohlhase2008} is perhaps most similar to Sophize Markdown but it needs creation of \LaTeX\space documents and use of \LaTeX ML systems. This makes its use in real-time web workflows impractical.

>>>>

\newcommand\todo[1]{\textcolor{red}{TODO: #1}}
\graphicspath {{../res}}
\begin{document}
%
\title{Sophize Mathematics Library}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Abhishek Chugh\inst{1}\orcidID{0000-0001-6765-5229}}
%
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Sophize Foundation, Bengaluru - 560066, India 
\email{abc@sophize.org}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Sophize is a novel mathematics library and discussion platform with a mission to help our
users find and organize mathematical proofs. We present Sophize's novel yet familiar
knowledge organization scheme that is used to represent a wide variety of proofs. With this
scheme at its core, we have engineered a platform to logically aggregate knowledge from
multiple sources. The platform curates knowledge from published research, encyclopaedias
such as PlanetMath and Wikipedia, informal computer programs, and formal systems such as
Metamath; and allows its users to run a federated search over them. To make this happen, two
developments were necessary. First, we came up with the concept of `proof-generating
machines' and developed an open plugin architecture that allows proofs from computer
programs to be generated as required. Second, we extended the Markdown language to represent
the connections between mathematical objects found across various sources of knowledge. In
addition, we also utilized this new language to create a novel communication system built
specifically to aid mathematicians in solving problems collaboratively.

@@

In the past couple of decades, the Internet has revolutionized our life by giving us access
to seemingly unlimited information. Information from a wide variety of topics ranging from
human medicine to rocket science is now just a few keystrokes away. However, it seems
that we are having even more trouble making sense of the World around us. In fact, dealing with the vast amounts of information thrown at us seems to be one of the most important problems of our generation. While we have access to a wide range of opinions on almost every topic imaginable, it has become increasingly difficult to decide which ones to trust. It seems that the Internet can find us "proofs" for anything we would like to believe.

Thus, the author's primary motivation for this project is to create a library that not just lays out all the information about the World but organizes it in a way that helps us make better sense of it. A library that helps us access justifications from a variety of
viewpoints and allows us to collaboratively and objectively evaluate issues in the
justifications or the viewpoints they arise from. Clearly, this is a challenging goal.

@@

Apart from organizing knowledge focused on trust and aggregating disparate data sources, logical organization of mathematics can lead to many other advantages. We list some interesting ones below.

@@

Another use case would be to understand the consequences of important conjectures like the Reimann hypothesis by creating two belief sets - one where the conjecture is true and one where it is false. The proof graphs would systematically store such conjectural knowledge which often gets lost especially when results from multiple conjectures are involved. Complexity theory is one such field where this need is quite apparent. Such conjectural knowledge knowledge can have practical applications too. For example, a lot of our information security systems utilize the knowledge developed with the conjectures $P \neq NP$.

Proof graphs are currently organized as trees as seen in Fig \ref{proof_tree}. Users can keep exploring the proof till they reach a point where the propositions are already familiar to them. A future application of proof graphs would be to generate narrative proofs tailored to the reader's current understanding. This could be achieved in a classroom setting where the current level of mathematical understanding of students is known or assumed. It may also be possible to infer this from the user's activity on the platform.

@@

\section{Proof-Generating Machines (PGM)}
\label{sec:pgm}
A lot of mathematical knowledge is computed. Most computation machines (e.g. calculators, algebras) can be seen as performing the following tasks
\begin{itemize}
\item Parsing the input into a language that can represent math formulae. 
\item Finding and outputting an equivalent form of the formula.
\end{itemize}
Thus, their focus is not on finding the proof but only finding the equivalent simplified representation of the input formula (Although many computer algebras do show the steps of the computation). With Sophize, the focus instead is to generate proofs of the input propositions. A \textbf{proof-generating machine} (PGM) is a computer program that generates a partial proof graph against or in favour of an argument. PGMs can also have the capability to perform tasks 1 and 2, mentioned above.

PGMs take in some text as input which is parsed by the each PGM as they wish. Typically the text is treated either as a proposition (e.g. `3 + 5 = 8') or a formula (eg. `5 + 6') and in the latter case the PGM is responsible for finding an appropriate proposition (eg. `5 + 6 = 11') for the given input. The PGM then generates a proof graph for this proposition and returns back to the user.

@@

\begin{enumerate}
\item Arguments whose the premises are limited to the premises of the PGM.
\item Special `Arguments' that indicate that the generation of the proof of this argument's conclusion must be delegated to another PGM. Seeing such arguments, the Sophize platform requests the remainder of the proof from the indicated PGMs as needed to allow users to browse the proof uninterrupted. To paginate large proofs, a PGM may also delegate to itself after generating some (say 100) arguments.
\end{enumerate}

\subsection{Materializing results}

@@

In the above example, the `EXPAND' option expanded the definition of the term, and statement of the proposition in place. Thus we were able to easily re-use concepts and theorems already extracted from other sources (say, Wikipedia and the Oxford dictionary in this case). Clicking on the underlined concepts pops-up a modal dialog with the definition of the term clicked.

@@

Sophize divides its data into groups called datasets for copyright and data access management. For example, the `wiki' dataset that extracts knowledge from Wikipedia mathematics glossaries releases its data under `CC-BY-SA' license, whereas the `metamath' dataset is public (`CC0'). Our signed-in users are free to create their own datasets and we allow them to choose who can read, write or comment in these datasets. Another interesting access control is the restriction of use of resources (especially argument-validity criteria) from an otherwise open-to-read dataset.

Information can be added to Sophize platform using intuitive web forms. However, it is infeasible to extract information from large repositories in this manner. Sophize publishes its data schema (explained in Section \ref{sec:core}) and helper libraries in multiple languages. The extracted information is can thus be stored as text (JSON) files in a repository. Sophize provides dataset owners the ability to sync knowledge such repositories to the central Sophize database. The datasets `wiki' and `planetmath' use this scheme for keeping the script-generated data synced with Sophize's central database.

\subsection{Search Interface}
\label{sec:search}
Sophize allows its users to run a run a federated search over its resources using an Elasticsearch server. We even allow users to \LaTeX to their search queries. The server runs search a space-insensitive and case-insensitive search over \LaTeX\space formulae by encoding \LaTeX\space formulae as lexemes and an Elasticsearch analyzer processes them appropriately. (See section \ref{sec:sources} for sources)

@@

We show how proof-generating machines can be used to instantaneously compute proofs that embed into previously computed graphs. We showcase how Sophize Markdown allow users to effortlessly browse proof-graphs and lookup entities in math literature. While there are other ways to represent semantic knowledge, such as content MathML and OpenMath, their scope is somewhat limited to specifying the meaning of the mathematical formula. The sTeX system \cite{Kohlhase2008} is perhaps most similar to Sophize Markdown but it needs creation of \LaTeX\space documents and use of \LaTeX ML systems. This makes its use in real-time web workflows impractical.
